Quantifying the importance of an inference model in Bayesian phylogenetics

### 2020-05-29 pirouette_article.tex
2nd hour

The development of new powerful Bayesian phylogenetic inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}] 
has been a major advance in constructing phylogenetic trees 
from character data (usually nucleotide sequences) extracted 
from extant (but also extinct, where possible) organisms, and hence in our understanding of the main drivers 
and modes of diversification.

@@

For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rate
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree}].
Despite such a rich theoretical landscape, only a few of these 
diversification models are available as tree priors in BEAST2.

@@

The first step ($T_1$) of the twinning process creates a tree $\tau_{\mathit{T}}$
with branching times $\Vec{t}_{\mathit{T}}$ while preserving the original
topology $\psi_{\mathit{G}}$:
\begin{align}
  \tau_{\mathit{G}} = (\Vec{t}_{\mathit{G}}, \psi_{\mathit{G}}) 
  \xrightarrow[]{\mathit{T_1}} 
  \tau_{\mathit{T}} = (\Vec{t}_{\mathit{T}}, \psi_{\mathit{G}})
\end{align}
The default option for the diversification model $p_T$ is the standard BD model.
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find 
the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS 
package [\cite{TESS, hohna2016tess}].

>>>>

The development of new powerful Bayesian phylogenetic inference tools, 
such as BEAST [\cite{drummond2007beast}], 
MrBayes [\cite{huelsenbeck2001mrbayes}]
or RevBayes [\cite{hohna2016revbayes}] 
has been a major advance in constructing phylogenetic trees 
from character data (usually nucleotide sequences) extracted
% 00004
from organisms (usually extant, but extinction events 
and/or time-stamped data can be added), 
and hence in our understanding of the main drivers 
and modes of diversification \richel{Reword this awkwardly long sentence}.

@@

For example, one can add novel diversification models 
by writing a BEAST2 package that contains the likelihood 
formula of a phylogeny under the novel diversification model, 
i.e. the prior probability of a species tree.
Many diversification models (and their associated probability algorithms) 
have been developed, e.g., models in which diversification is time-dependent [\cite{nee1994reconstructed,rabosky2008explosive}],
or diversity-dependent [\cite{etienne2012diversity}],
or where diversification rates change for specific lineages 
and their descendants [\cite{etienne2012conceptual, rabosky2014automatic, alfaro2009nine}].
Other models treat speciation as a process that takes 
time [\cite{rosindell2010protracted, etienne2012prolonging, lambert2015reconstructed}],
or where diversification rate
depends on one or more traits [\cite{maddison2007estimating, fitzjohn2012diversitree}].
Despite such a rich theoretical landscape, only a few of these 
diversification models are available as tree priors in BEAST2
% 00005
\richel{
  This is awkward: most of these models are available as plug-ins,
  which was the whole idea when making BEAST2. This is a great moment
  to mention the tree models available in those BEAST2 packages
}
.

@@

The first step ($T_1$) of the twinning process creates a tree $\tau_{\mathit{T}}$
with branching times $\Vec{t}_{\mathit{T}}$ while preserving the original
topology $\psi_{\mathit{G}}$:
\begin{align}
  \tau_{\mathit{G}} = (\Vec{t}_{\mathit{G}}, \psi_{\mathit{G}}) 
  \xrightarrow[]{\mathit{T_1}} 
  \tau_{\mathit{T}} = (\Vec{t}_{\mathit{T}}, \psi_{\mathit{G}})
\end{align}

\new{
  % 00003
  The default option for the twin diversification model $p_T$ 
  is to use the standard BD model.
}
It is then possible to use the likelihood function 
$L_{\mathit{T}}$ for this diversification model to find 
the parameters $\theta^{*}_{\mathit{T}}$ 
(e.g. speciation and extinction rates, in case of a BD model) 
that maximize this likelihood applied 
to the true tree, conditioned on its number of tips $n_{\mathit{G}}$:
\begin{align}
    \max[L_{\mathit{T}}(\theta_{\mathit{T}}|\tau_{\mathit{G}}, n_{\mathit{G}})] 
\rightarrow \theta^{*}_{\mathit{T}}.
\end{align}
We use $\theta^{*}_{\mathit{T}}$ to simulate a number 
$n_{\mathit{T}} = n_{\mathit{G}}$ 
of branching times $\Vec{t}_{\mathit{T}}$ for the twin tree 
$\tau_{\mathit{T}}$, under the process $p_{T}$, 
while preserving the topology. 
We simulate the new branching times using the TESS 
package [\cite{TESS, hohna2016tess}].
